FITS format for planetary surfaces: definitions, applications and best practices.

### 2018-05-11 2018EA000388_Answers_1Reviewer.txt

l. 175
Rephrased: see the new version of the manuscript

l. 177
CUNIT is not mandatory in FITS standard: when absent, degrees are assumed.
This is a valid assumption for planetary surfaces too. 
Rephrased: see the new version of the manuscript

Table 2 : header
Mercury must be there... sorry for the oversight.
TNO and KBO are Asteroids: this is now clarified in the note.

l. 257
ORIGIN?

l.275-277
Images have been modified to better show the equivalence of georeferenced TIFF or FITS.
The paragraph has been modified accordingly.

l. 279
I have added a footnote to point the reader to a virtual header tutorial, in order to not to charge the
text with overwhelming technical details.

l. 307
Fixed

l. 347
I have added a paragraph: GDAL implentation is in progress.

l. 351
Rephrased

l. 389
Rephrased

l. 402
Rephrased

l. 404
Rphrased

### 2018-05-11 ess_marmo.tex
Almost done, only ORIGIN translation is missing.

\subsubsection{Visualization of Global and Projected Maps}
\label{sssec:maps}
High level products are meant to be used for visual inspection or GIS analysis.
They are distributed as projected rasters, often in
GeoTIFF\footnote{http://geotiff.osgeo.org/}
format, as GeoTIFF make them GIS ready.
Given some new keywords introduced in FITS and WCS definitions adapted to planetary
surface, FITS and GeoTIFF are completely interchangeable.
As an example we have downloaded some global maps from USGS
PDS Annex\footnote{https://astrogeology.usgs.gov/search} and converted them in FITS.
We used ds9 to display the FITS image and were able to visualize coordinates in meters
as in QGIS, a popular open source GIS software, but also in planetary latitude and
longitude thanks to the alternate coordinate
description (see figure \ref{fig:ds9qgis}, left side).
\begin{figure}[ht!]
\centerline{\includegraphics[height=17pc]{ess_marmo2} \includegraphics[height=17pc]{ess_marmo1}}
\caption{left: Mercury South Pole FITS image displayed in ds9;
right: GeoTIFF raster (displayed in yellow) and the virtual
header linked to the FITS raster (displayed in red); the two layers
are perfectly superposed.
Mercury image credits: MESSENGER Team, Arizona State University, Johns Hopkins Applied
Physics Laboratory, Carnegie Science}
\label{fig:ds9qgis}
\end{figure}

Assuming projected products are distributed in FITS format, a quick method to add GIS
support for FITS images is to use a GDAL-supported detached header.
GDAL Virtual Header files simply describe the internal structure of a FITS image, but
also the map projection and body size in a standardized "well known text"
(WKT) projection string.
As a result, the FITS raster loaded into QGIS thanks to its corresponding GDAL Virtual
Header is indistinguishable from the GeoTIFF layer (see figure \ref{fig:ds9qgis},
right side).

@@

As an example of non projected planetary data we have converted in FITS format some data
acquired by the imaging spectrometer OMEGA on-board of Mars Express.
OMEGA unprojected and uncalibrated data (level1b) are distributed by ESA-PSA and PDS
archives.
Their structure is described in the OMEGA Experiment Archive 
Interface Control Document\footnote{ftp://psa.esac.esa.int/pub/mirror/MARS-EXPRESS/OMEGA/MEX-M-OMEGA-2-EDR-FLIGHT-V1.0/DOCUMENT/EAICD\_OMEGA.PDF}.
\begin{figure}[ht!]
\centerline{\includegraphics[height=17pc]{ess_marmo4}\hspace{2pc}\includegraphics[height=17pc]{ess_marmo3}\hspace{2pc}\includegraphics[height=17pc]{ess_marmo5}}
\caption{left: OMEGA data structure (from OMEGA EAICD), science data and geometry
description are distributed in different files; center: FITS Multi Digital Object
scheme; as an example, we stored the spectral data in the first cube extension,
the coordinate look-up table in the second extension, and other geometry parameters
(incidence and emission angle, etc.) as images in the
following extensions; right: BSQ cube visualization with ds9.}
\label{fig:omegascheme}
\end{figure}

@@

\subsection{FITS and QGIS}
QGIS is a largely used open source geospatial analysis software.
It is using GDAL for layer import: QGIS is then already capable to display FITS files,
making use of a Virtual GDAL header, as described above.
Once implemented in GDAL, FITS will officially become a supported format by QGIS.

In addition to that, a QGIS plugin has been developed\footnote{https://github.com/epn-vespa/VO\_QGIS\_plugin}
\citep[][in preparation]{Minin} in order to download and display files exposed via EPN-TAP servers.

\subsection{MATISSE}
MATISSE\footnote{https://tools.asdc.asi.it/matisse.jsp} (Multi-purpose Advanced Tool
for the Instruments for the Solar System Exploration) \citep{ZINZI2016} is the webtool
designed by the Space Science Data Center of the Italian Space Agency (SSDC-ASI) to
access, visualize and analyze data from solar system exploration missions.
Currently, it provides access to public data from five different targets (i.e., Mercury,
Venus, Mars, dwarf planet Ceres and asteroid Vesta) using both data stored on SSDC local
storage or by querying remote dataset, such as VESPA, Planet Server and NASA JPL and
can be integrated with other datasets thanks to its modular structure.
One of th MATISSE 2D outputs is formatted as FITS, projected by interpolating the
original observations to a grid regularly spaced
in planetocentric coordinates (i.e., latitude and longitude).
We used the planetary FITS standard described in this work in order to make the output
file compliant with most FITS readers, especially DS9 and JS9.
The files generated by MATISSE could be also opened with standard GIS software (e.g.,
QGIS, ArcGIS) 
after being provided as input of the fits2vrt Python library\footnote{https://github.com/epn-vespa/fits2vrt},
whose output is a VRT virtual GDAL header pointing to the FITS file with the appropriate
projection.
Thanks to this approach, in the next major release of MATISSE (v2.0, planned to be
online by the end of 2018) we are planning to make use of JS9 as the standard reader for
the 2D online output of the tool, thus allowing users to perform advanced analysis in
both 2D and 3D directly inside a web-based environment.

\subsection{Aladin}
Aladin is both a sky atlas and a portal to access data -- mainly images and
catalogues -- available through the VO.
It is able to load FITS files and to display images taking into account various
WCS projections.
Like DS9, Aladin is popular and broadly used in the part of the astronomical
community manipulating images.
So far it is also the reference generator and visualizer of Hierarchical
Progressive Survey (HiPS).
HiPS\footnote{http://www.ivoa.net/documents/HiPS/} is a recommendation endorsed
by the International Virtual Observatory Alliance (IVOA).
It describes how to access to sky survey data stored at various spatial
resolutions, offering a progressive view of possibly very large datasets.
HiPS are usually generated from a collection of WCS tagged FITS files.
Aladin\footnote{http://aladin.u-strasbg.fr/} \citep{aladin} and its HiPS generator have
been updated to support planetary conventions (see Figure \ref{fig:aladin}).

@@

In the future, Aladin will be able to recognize the FITS keywords defined
in the present paper (see Tab. \ref{table:proj} and \ref{table:ctype}).
For example, it will be able to take into account a planet's radii to compute
distances on the surface of the body.
Finally, among the WCS projections listed in Tab. \ref{table:proj}, several of
them are barely used in extrasolar astronomy and are not supported in Aladin yet.
An effort has been started to revamp and update the code of projections in
Aladin.
We expect all projections to be available in a forthcoming release.

@@

FITS as an open and flexible data and metadata exchange and archiving format, could be
inserted in the PDS4 data model scheme in order to address the issue of more complex
metadata and data relationships.

In order to improve interoperability and knowledge sharing between astronomy and
planetary science, we will soon finalize the integration of FITS specifications in the
GDAL library.
In the framework of the VESPA project a review of the available and planned mission data
archives is ongoing in order to evaluate their insertion in the planetary VO structure
with a possible on-demand conversion in FITS format.

@@

\begin{thebibliography}{20}
\providecommand{\natexlab}[1]{#1}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
  \urlstyle{rm}\Url}\fi

@@

>>>>

\subsubsection{Visualization of Global and Projected Maps}
\label{sssec:maps}
High level products are meant to be used for visual inspection or GIS analysis.
They are distributed as projected rasters, often in
GeoTIFF\footnote{http://geotiff.osgeo.org/}
format, as GeoTIFF make them GIS ready.
Given some new keywords introduced in FITS and WCS definitions adapted to planetary
surface, FITS and GeoTIFF are completely interchangeable.
As an example we have downloaded a Mercury polar map from the USGS
PDS Annex\footnote{https://astrogeology.usgs.gov/search} and converted it in FITS.
The FITS image has been cropped in two smaller versions using the open source software
SWarp\footnote{https://www.astromatic.net/software/swarp} \citep{swarp}: SWarp was able to
automatically manage WCS planetary information.
We used ds9 to display the three FITS images and were able to correctly align them.
Linear coordinates in meters are visualized (see figure \ref{fig:ds9qgis}, left side)
as in QGIS, a popular open source GIS software, thanks to the alternate coordinate
description.
\begin{figure}[ht!]
\centerline{\includegraphics[height=17pc]{ess_marmo2} \includegraphics[height=17pc]{ess_marmo1}}
\caption{left: Mercury South Pole FITS image displayed in ds9, alternate coordinates in meters
are displayed;
right: GeoTIFF raster (displayed in red) and the two virtual haeders
linked to FITS rasters (displayed in green and blue); the three layers
are perfectly superposed.
Mercury image credits: MESSENGER Team, Arizona State University, Johns Hopkins Applied
Physics Laboratory, Carnegie Science}
\label{fig:ds9qgis}
\end{figure}

Assuming projected products are distributed in FITS format, a quick method to add GIS
support for FITS images is to use a GDAL-supported detached header.
GDAL Virtual Header\footnote{http://www.gdal.org/gdal\_vrttut.html}
files simply describe the internal structure of a FITS image, but
also the map projection and body size in a standardized "well known text"
(WKT) projection string.
As a result, the FITS raster loaded into QGIS thanks to its corresponding GDAL Virtual
Header is indistinguishable from the GeoTIFF layer (see figure \ref{fig:ds9qgis},
right side).

@@

As an example of non projected planetary data we have converted in FITS format some data
acquired by the imaging spectrometer OMEGA on-board of Mars Express.
OMEGA unprojected and uncalibrated data (level1b) are distributed by ESA-PSA \citep{BESSE2018131}.
Their structure is described in the OMEGA Experiment Archive 
Interface Control Document\footnote{ftp://psa.esac.esa.int/pub/mirror/MARS-EXPRESS/OMEGA/MEX-M-OMEGA-2-EDR-FLIGHT-V1.0/DOCUMENT/EAICD\_OMEGA.PDF}.
\begin{figure}[ht!]
\centerline{\includegraphics[height=17pc]{ess_marmo4}\hspace{2pc}\includegraphics[height=17pc]{ess_marmo3}\hspace{2pc}\includegraphics[height=17pc]{ess_marmo5}}
\caption{left: OMEGA data structure (from OMEGA EAICD), science data and geometry
description are distributed in different files; center: FITS Multi Digital Object
scheme; as an example, we stored the spectral data in the first cube extension,
the coordinate look-up table in the second extension, and other geometry parameters
(incidence and emission angle, etc.) as images in the
following extensions; right: BSQ cube visualization with ds9.}
\label{fig:omegascheme}
\end{figure}

@@

\subsection{FITS and QGIS}
QGIS is a largely used open source geospatial analysis software.
It is using GDAL for layer import: QGIS is then already capable to display FITS files,
making use of a Virtual GDAL header, as described above.
The complete implementation of FITS metadata in GDAL library is in progress.
Once implemented in GDAL, FITS will officially become a supported format by QGIS.

In addition to that, a QGIS plugin has been developed\footnote{https://github.com/epn-vespa/VO\_QGIS\_plugin}
\citep{Minin} in order to download and display files exposed via EPN-TAP servers.

\subsection{MATISSE}
MATISSE\footnote{https://tools.asdc.asi.it/matisse.jsp} (Multi-purpose Advanced Tool
for the Instruments for the Solar System Exploration) \citep{ZINZI2016} is the webtool
designed by the Space Science Data Center of the Italian Space Agency (SSDC-ASI) to
access, visualize and analyze data from solar system exploration missions.
Currently, MATISSE provides access to public data from five different targets (i.e., Mercury,
Venus, Mars, the dwarf planet Ceres and the asteroid Vesta).
MATISSE 2D outputs are formatted as FITS, in equirectangular projection and in planetocentric coordinates.
The planetary FITS standard described in this work has been implemented in MATISSE in order to make the output
file compliant with most FITS readers, especially DS9 and JS9.
The files generated by MATISSE could be also opened with standard GIS software (e.g.,
QGIS, ArcGIS) 
after being provided as input of the fits2vrt Python library\footnote{https://github.com/epn-vespa/fits2vrt},
whose output is a VRT virtual GDAL header pointing to the FITS file with the appropriate
projection.
Thanks to this approach, in the next major release of MATISSE (v2.0, planned to be
online by the end of 2018) JS9 will be the standard reader for
the 2D online output of the tool, thus allowing users to perform advanced analysis in
both 2D and 3D directly inside a web-based environment.

\subsection{Aladin}
Aladin is both a sky atlas and a portal to access data, mainly images and
catalogues, available through the VO.
Aladin is able to load FITS files and to display images taking into account various
WCS projections.
So far it is also the reference generator and visualizer of Hierarchical
Progressive Survey (HiPS).
HiPS\footnote{http://www.ivoa.net/documents/HiPS/} is a recommendation endorsed
by the International Virtual Observatory Alliance (IVOA).
It describes how to access to sky survey data stored at various spatial
resolutions, offering a progressive view of possibly very large datasets.
HiPS are usually generated from a collection of WCS tagged FITS files.
Aladin\footnote{http://aladin.u-strasbg.fr/} \citep{aladin} and its HiPS generator have
been updated to support planetary conventions (see Figure \ref{fig:aladin}).

@@

In the future, Aladin will be able to recognize the FITS keywords defined
in the present paper (see Tab. \ref{table:proj} and \ref{table:ctype}).
Finally, among the WCS projections listed in Tab. \ref{table:proj}, several of
them are barely used in extrasolar astronomy and are not supported in Aladin yet.
An effort has been started to revamp and update the code of projections in
Aladin.

@@

FITS is an open and flexible data and metadata exchange and archiving format, its use
in the PDS4 data model scheme would solve the issue of complex
metadata and data relationships separating data processing and the data archiving processes.

In the framework of the VESPA project a review of the available and planned mission data
archives is ongoing in order to evaluate their insertion in the planetary VO structure
with a possible on-demand conversion in FITS format.
The integration of FITS specifications in the GDAL library will be soon finalize with the goal to
improve data processing efficiency in planetary science.

@@

\begin{thebibliography}{22}
\providecommand{\natexlab}[1]{#1}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
  \urlstyle{rm}\Url}\fi

@@

\bibitem[{\textit{{Bertin} et~al.}(2002)\textit{{Bertin}, {Mellier},
  {Radovich}, {Missonnier}, {Didelon}, and {Morin}}}]{swarp}
{Bertin}, E., Y.~{Mellier}, M.~{Radovich}, G.~{Missonnier}, P.~{Didelon}, and
  B.~{Morin} (2002), {The TERAPIX Pipeline}, in \textit{Astronomical Data
  Analysis Software and Systems XI}, \textit{Astronomical Society of the
  Pacific Conference Series}, vol. 281, edited by D.~A. {Bohlender},
  D.~{Durand}, and T.~H. {Handley}, p. 228.

\bibitem[{\textit{Besse et~al.}(2018)\textit{Besse, Vallat, Barthelemy, Coia,
  Costa, Marchi, Fraga, Grotheer, Heather, Lim, Martinez, Arviset, Barbarisi,
  Docasal, Macfarlane, Rios, Saiz, and Vallejo}}]{BESSE2018131}
Besse, S., C.~Vallat, M.~Barthelemy, D.~Coia, M.~Costa, G.~D. Marchi, D.~Fraga,
  E.~Grotheer, D.~Heather, T.~Lim, S.~Martinez, C.~Arviset, I.~Barbarisi,
  R.~Docasal, A.~Macfarlane, C.~Rios, J.~Saiz, and F.~Vallejo (2018), Esa's
  planetary science archive: Preserve and present reliable scientific data
  sets, \textit{Planetary and Space Science}, \textit{150}, 131 -- 140,
  \doi{https://doi.org/10.1016/j.pss.2017.07.013}, enabling Open and
  Interoperable Access to Planetary Science and Heliophysics Databases and
  Tools.
